---
title: "Introduction to hydroGraphR"
author: "Paula Soto<br>"
date:  "`r Sys.Date()`"
output:
  github_document:
    html_preview: true
always_allow_html: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

`hydroGraphR` is an R package designed to streamline the process of visualizing hydrometric data by generating hydrographs that compare historical water level or discharge statistics with a specific year's observed data. The package is particularly useful for hydrologists, environmental scientists, and water resource managers who need to analyze trends, seasonal variations, and anomalies in hydrological data.

This vignette provides an overview of the package’s key functionalities and demonstrates its application in a typical workflow. Users can leverage `hydroGraphR` to:

- Load and preprocess hydrometric data: The package supports various data formats and includes functions for cleaning, filtering, and formatting time series data.
- Compute historical statistics: It can calculate long-term percentiles, mean values, and other statistical summaries to establish a historical baseline for comparison.
- Generate hydrographs: Users can create clear and customizable plots that overlay the current year’s hydrometric data on historical trends, making it easy to identify deviations or anomalies.
- Customize visualization settings: The package allows for fine-tuned control over axis scaling, line styles, colors, and annotations, ensuring clarity and effectiveness in communication.

By automating many of the repetitive tasks involved in hydrograph creation, hydroGraphR enhances efficiency and consistency in hydrological data analysis. The following sections will guide you through its core functionality with step-by-step examples.

## Installation

To install the development version of `hydroGraphR` from GitHub:


```{r setup, eval = FALSE}
# Install devtools if not already installed
install.packages("devtools")

library(devtools)

# Install hydroGraphR
install_github("pausoto7/hydroGraphR")

# Download library
library(hydroGraphR)


```

```{r dl package, echo=FALSE}

#actually run library
library(hydroGraphR)

```

## Workflow Overview

A typical workflow for hydroGraphR involves the following steps:

1. Download hydrometric data for specific WSC stations.
2. Generate single-year and historical statistics.
3. Visualize the data by creating hydrographs.

### Step 1: Download Hydrometric Data

Use the `dl_hydro()` function to download data for specific Water Survey of Canada (WSC) station numbers. You can find station numbers on the  [WSC website](https://wateroffice.ec.gc.ca/search/real_time_e.html). Note that with this function historical and real time data will be downloaded. Typically, historical data has been QAQC'd and occurred longer than two years ago. Real time data is data collected within the last two years and **has not** been QAQC'd. 

You may also assign nick names to stations for easier identification in subsequent analyses. If you are going to use nick names ensure that the number of nick names matches the number of station numbers. 

Examples:

```{r dl}
#example of station(s) download without nicknames
hydro_data <- dl_hydro(station_number = c("08LD001", "08LC002"))

#example of station(s) download with nick names. 
hydro_data_nickname <- dl_hydro(station_number = "08LD001", nick_name = "Adams River")
```

### Step 2: Create Statistics for Single Year and Historical Data

Once the hydrometric data is downloaded, you can calculate statistics for: 

- A specific year of interest (YOI) using `create_hydro_stats_singleYr()`.
- Historical data using `create_hydro_stats_historical()`. 

Statistics calculated are listed in more detail in Step 3. 

WY is a logical value indicating whether to present hydrograph by water year (Nov-Oct) (TRUE) or calendar year (Jan-Dec) (FALSE).

Max/min dates can also be selected for historical dates if focus is on a specific period.For example, you could enter date_minimum =  "2010-01-01",  date_maximum - "2019-12-31" and YOI = 2020 if you wanted to compare 2020 with the 2010's. 

***Important Note: Data from the past two years may be provisional, as such should be used with caution when a recent YOI is selected. Status of data can be found on the WSC website.***

<br>


```{r getstats}
all_hydro_sites_1yr <- create_hydro_stats_singleYr(hydro_data, 
                                                   YOI = 2021, 
                                                   WY = FALSE) # Calendar year

all_hydro_sites_hist <- create_hydro_stats_historical(hydro_data) # Use all available hydrometric data


```

### Step 3: Create  Hydrographs

Visualize your data using one of two functions, depending on your preferred output style. 

- `create_hydrograph_separate()`- Generates individual hydrographs for each station. 

- `create_hydrograph_faceted()` - Creates a single faceted plot displaying multiple stations together.

Variables shown on hydrograph legends are described here: 

```{r def table, echo=FALSE, results='asis'}

definitions <- data.frame(
  Term = c("Mean",  "Mean ", 
            "Q25 (Today)", "Q75 (Today)", "MAD",
           "Date Ranges"),
  
  Definition = c("The average value of the measurements over the YOI period.", 
                 "The average value of the measurements over the historical period.",
    "The 25th percentile of today's data distribution, indicating that 25% of the data points are below this value.",
    "The 75th percentile of today's data distribution, indicating that 75% of the data points are below this value.",
    "Mean Annual Discharge, a measure of the average amount of discharge through a river or stream over the course of a year.",
    "The years during which the data was collected and included in analysis."
  )
)

knitr::kable(definitions, "html", escape = FALSE, col.names = c("Term", "Definition")) %>%
  kableExtra::kable_styling(full_width = TRUE, position = "left", bootstrap_options = "condensed") %>%
  kableExtra::pack_rows(
    index = c("YOI"  = 1, 
              "Historical" = 3)
  )
  


```

#### Variables

- **Parameter**: Options are `"flow"` for a discharge hydrograph, and `"level"` for a water level hydrograph. 

- **WY**: Logical value indicating whether to present hydrograph by water year (Nov-Oct)(`TRUE`) or calendar year (Jan-Dec)(`FALSE`).
    - No matter the output selection ensure that the water year (WY) choice matches the selection made in Step 2. For example, if "calendar year" was chosen for statistics in Step 2, it must also be selected for the hydrograph presentation. Selecting a different year type will result in missing data on the hydrograph and trigger a warning.

- **output_type**: 
    - `"print"` will print your image in R, useful for embedding in rmarkdown or shiny type outputs. 
    - `"jpeg"` will produce a jpeg image and save it to the "figures/" folder of this project. 

<br>

#### **Option 1**: `create_hydrograph_separate()`

- This function creates separate hydrographs as individual images which are ideal for standalone use or printing. When the "jpeg" option is selected in this option, an additional title will appear on the image with the station name to avoid confusion of station ID if the file name is not easily available (for example if the image is pasted into a word document). 

```{r hydrographsep, fig.width=7, fig.height=4}

create_hydrograph_separate(
  all_hydro_sites_hist,
  all_hydro_sites_1yr,
  parameter = "flow", # Discharge hydrograph
  output_type = "print",
  WY = FALSE # calendar year
)

```
<br><br><br>

#### **Option 2**: `create_hydrograph_faceted()`

- This function creates a single faceted hydrograph, making it easy to compare multiple stations side by side. There are a few additional variabels that can also be modified in this function: 

    - `fixed_y_scales`: A character string specifying whether the y-axis scale is "fixed" or "free" across facets. Defaults to "fixed".
    - `custom_ymax_input`: A numeric value for a custom maximum y-axis value. Leave as NA for automatic ymax.
    - `custom_ymin_input`:A numeric value for a custom minimum y-axis value. Leave as NA for automatic ymin.
    - `jpeg_width`: A numeric value specifying the width of the figure (in inches) for JPEG output. 6 inches is automatic output.  
    - `jpeg_height`: A numeric value specifying the height of the figure (in inches) for JPEG output. 8 inches is automatic output. 


```{r hydrofacet, fig.width=7, fig.height=7.5}

create_hydrograph_faceted(
  all_hydro_sites_hist,
  all_hydro_sites_1yr,
  parameter = "flow",
  WY = FALSE, # calendar year
  output_type = "print", 
  fixed_y_scales = "fixed",
  custom_ymax_input = NA, 
  custom_ymin_input = NA
)
```


### Additional tools

You can combine these tools with other R packages to create streamlined workflows for hydrologic data analysis. For example, the `streamTrackR` package complements `hydroGraphR` by enabling easy comparison of current conditions across selected rivers.


### Summary

`hydroGraphR` simplifies hydrometric data analysis and visualization with functions for downloading data, generating statistics, and creating hydrographs. Whether working with individual rivers or a larger dataset, the package offers flexibility for both exploratory and presentation-ready outputs.

For further details, consult the package documentation
